Hypertension and microbiome

### 2019-05-16 articletwo.Rmd
Cleaning salt results

### Lactobacillus

Load genus data with dU-NA and combine it in a data frame with clr transformed lactobacillus abundances.

@@

### Daily urine natrium

Associations between lactobacillus and dU-NA

```{r abund salt lm continuous}
lm.lacto.duna <- lm(lactobacillus.bacteria ~ dUNA + BL_AGE + SEX, data = df.genus.salt)
lm.lacto.duna %>% tidy %>% kable
```

@@

@@

### Salt groups

Association between lactobacillus and salt groups

```{r abund salt lm group}
lm.lacto.saltgroup <- lm(lactobacillus.bacteria ~ salt_groups + BL_AGE + SEX, data = df.genus.salt)
lm.lacto.saltgroup %>% tidy %>% kable
```

Assesment for linear model asuumptions

@@

Anova between lactobacillus and salt groups

@@

Tukey's range test for lactobacillus and salt groups

@@

### Adjusted means

Calculate adjusted means for lactobacillus

@@

### Odds for hypertension

Odds for hypertension

```{r glm htn lacto}
glm.hypertension.lactogroup <- glm(HYPERTENSION ~ lacto_groups + BL_AGE + SEX,
                  family = binomial(link=logit),
                  data = df.genus.salt) %>%
    broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE)
glm.hypertension.lactogroup %>% kable
```

```{r odds htn lacto}
odds.hypertension.lactogroup <- glm.hypertension.lactogroup %>%
    filter(!grepl("Intercept", term)) %>%
    select(term, estimate, conf.low, conf.high) %>%
    add_row(term = "lacto_groups1", estimate = 1.0, conf.low = NA, conf.high = NA) %>%
    dplyr::filter(grepl("lacto", term))
```

### Twin-forest plot

>>>>

## Lactobacillus (figure 4)

The fourth and final step is to study how the amount of dietary salt
affect the lactobacillus abundances and how these abundances affect
the odds for hypertension. We load genus data with dU-NA and combine
it in a data frame with CLR transformed lactobacillus abundances.

@@

Characteristics in salt analysis. Analyses are limited by the number
of participants with collected dU-NA.

```{r salt characteristics}
pseq.genus.salt %>%
    meta %>%
    gather(key,value) %>%
    group_by(key) %>%
    summarise_all(funs(N = n(),
                       Mean = mean(as.numeric(value), na.rm = TRUE),
                       sd = sd(as.numeric(value), na.rm = TRUE),
                       NAs = sum(is.na(value))))  %>%
    mutate_if(is.numeric, round, 2) %>%
    kable()
```

### How sodium affects lactobacillus abundances?

Associations between continuous dU-NA and lactobacillus abundances.

```{r abund salt lm continuous}
lm.lacto.duna <- lm(lactobacillus.bacteria ~ dUNA + BL_AGE + SEX, data = df.genus.salt)
lm.lacto.duna %>% tidy %>% mutate_if(is.numeric, round, 3) %>% kable
```

Testing assumptions for linear regression analysis

@@

Diagnostic plots for linear model

@@

### Is there associations between salt groups and abudances?

Associations between categorical dU-NA split in three equal size groups and lactobacillus abundances

```{r abund salt lm group}
lm.lacto.saltgroup <- lm(lactobacillus.bacteria ~ salt_groups + BL_AGE + SEX, data = df.genus.salt)
lm.lacto.saltgroup %>% tidy %>% mutate_if(is.numeric, round, 3) %>% kable
```

Testing assumptions for linear regression analysis

@@

Bartlett's test for equality of variance gives p = `r
bartlett.test(lactobacillus.bacteria ~ salt_groups, data =
df.genus.salt)$p.value`.

One-way test assuming normal distributions gives for equal mean p = `r
oneway.test(lactobacillus.bacteria ~ salt_groups, data =
df.genus.salt, var.equal = TRUE)$p.value`.

Analysing also differences between salt groups using ANOVA.

@@

Tukey's range for pairwise test between salt groups

@@

### Adjusting means for forest plot

Calculate adjusted means for lactobacillus assuming male sex and mean age.

@@

## Is there associations between lactobacillus abudance and hypertension?

Studying associations between blood pressure variables and lactobacillus abundances

```{r lactobacillus RR}
calculateglm(df.genus.salt,
             modelstring = "%s ~ lactobacillus.bacteria + BL_AGE + SEX",
             filterstr = "lactobacillus") %>% 
    select(response, term, estimate, std.error, p.value) %>%
    mutate_if(is.numeric, round, 3) %>%
    kable
```

Checking assumptions for linear regression for MAP

```{r table map lactobacillus}
gvlma.table.plot(lm(MAP ~ lactobacillus.bacteria + BL_AGE + SEX, data = df.genus.salt))$table %>% kable
```

Diagnostic plots for linear model

```{r plot map lactobacillus, fig.width=12, fig.height=14}
gvlma.table.plot(lm(MAP ~ lactobacillus.bacteria + BL_AGE + SEX, data = df.genus.salt))$plot %>% plot
```

```{r odds htn lacto}
odds.hypertension.lactogroup <- glm.hypertension.lactogroup %>%
    filter(!grepl("Intercept", term)) %>%
    select(term, estimate, conf.low, conf.high) %>%
    add_row(term = "lacto_groups1", estimate = 1.0, conf.low = NA, conf.high = NA) %>%
    dplyr::filter(grepl("lacto", term))
odds.hypertension.lactogroup %>% kable
```

Odds for hypertension by categorical lactobacillus abundance.

```{r glm htn lacto groups}
glm.hypertension.lactogroup <- glm(HYPERTENSION ~ lacto_groups + BL_AGE + SEX,
                  family = binomial(link=logit),
                  data = df.genus.salt) %>%
    broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>%

odds.hypertension.lactogroup <- glm.hypertension.lactogroup %>%
    filter(!grepl("Intercept", term)) %>%
    select(term, estimate, conf.low, conf.high) %>%
    add_row(term = "lacto_groups1", estimate = 1.0, conf.low = NA, conf.high = NA) %>%
    dplyr::filter(grepl("lacto", term))

glm.hypertension.lactogroup %>%
    mutate_if(is.numeric, round, 3) %>%
    kable
```

## Twin-forest plot
